---
title: "galton-height-regression"
output: html_document
---

# Predicting Child Height from Parental Height: A Linear Regression Analysis of the Galton Height Dataset

## Project Summary
Height is known to have a strong genetic component, and parental height is often used as a predictor of child height. In this analysis, we examined the relationship between midparent height (the average height of both parents) and child height using the Galton height dataset. A simple linear regression model was fitted to evaluate how well midparent height predicts child height. The results indicated a significant positive relationship between the two variables, with midparent height explaining approximately 10.3% of the variance in child height. These findings suggest that while parental height is an important factor, other genetic and environmental factors also contribute to a child's height.

## Introduction
Human height is influenced by a combination of genetic and environmental factors, but parental height is often considered a strong predictor of a child's height . Early statistical work by Francis Galton demonstrated that the average height of both parents, known as midparent height, provides a strong indicator of a child’s expected height (Galton, 1886)Galton also observed that children’s heights tend to regress toward the population average rather than exactly matching their parents, a phenomenon later known as regression toward the mean. The Galton height dataset, collected in the late 19th century, contains measurements of parents’ heights and the heights of their children. The dataset includes a variable called midparent height, which represents the average height of both parents, and allow us to study how parental height relates to the height of their child. In this analysis, we aim to answer the question: **Can a child’s height be predicted based on their parents’ heights?** To address this question, we use a simple linear regression model to examine the relationship between midparent height and child height. By analyzing this relationship, we can better understand how parental height contributes to variation in child height.

## Methods

To investigate the relationship between parental height and child height, we conducted a mutiple linear regression analysis using the Galton height dataset. The dataset was randomly split into training (80%) and testing (20%) sets. The training dataset was used to fit the regression model, while the testing dataset was used to evaluate the model's predictive performance.

A multiple linear regression model was fitted with child height as the response variable, and the midparent height (the average height of the mother and father) and child gender as predictor variables.

## Set-up and Reading in Data

```{r}
#importing packages and libraries 
library(tidyverse)
library(tidymodels)
library(ggplot2)


set.seed(123)
```

```{r}
#reading in the data

url <- "https://raw.githubusercontent.com/vincentarelbundock/Rdatasets/master/csv/HistData/GaltonFamilies.csv"
galton_raw <- readr::read_csv(url, show_col_types = FALSE)

GaltonFamilies <- galton_raw |>
  dplyr::select(-any_of(c("rownames", "X"))) |>
  janitor::clean_names() |>
  dplyr::mutate(gender = as.factor(gender))



## Exploratory Data Analysis

```{r}
head(GaltonFamilies)
str(GaltonFamilies)
summary(GaltonFamilies)
```
The GaltonFamilies dataset contains information about the heights of children and their parents from a number of families. Each row in the dataset represents one child, and there are 934 observations with 8 variables in total. The dataset includes variables such as the height of the father, the height of the mother, the calculated midparent height, the child’s gender, and the child’s height.

From the summary statistics, the average child height is about 66.7 inches, with heights ranging from 56 inches to 79 inches. The dataset also contains slightly more male children than female children. After checking for missing values, we found that there are no missing observations, meaning the dataset is complete and can be used directly for further analysis.

```{r}
colSums(is.na(GaltonFamilies))
```

```{r}

ggplot(GaltonFamilies, aes(x = child_height)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "black") +
  labs(
    title = "Distribution of Child Height",
    x = "Child Height (inches)",
    y = "Count"
  )
```
The histogram shows that child heights are roughly normally distributed. Most children in the dataset have heights between about 60 and 72 inches, with the center of the distribution around 66–67 inches. There are fewer observations at the very short and very tall ends of the distribution. This pattern suggests that child height follows a fairly typical bell-shaped distribution, which is common for biological traits like height.

```{r}
ggplot(GaltonFamilies, aes(x = midparent_height, y = child_height)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "red") +
  labs(
    title = "Child Height vs Midparent Height",
    x = "Midparent Height",
    y = "Child Height"
  )
```
This scatterplot explores the relationship between a child’s height and the average height of their parents (midparent height). There is a clear positive trend, meaning that children with taller parents tend to be taller themselves. Although there is still variability in the data, the upward trend suggests that parental height is an important factor influencing a child's height. This relationship indicates that midparent height may be a useful predictor when building a model to estimate child height.

```{r}
ggplot(GaltonFamilies, aes(x = gender, y = child_height)) +
  geom_boxplot(fill = "lightblue") +
  labs(
    title = "Child Height by Gender",
    x = "Gender",
    y = "Child Height"
  )
```

The boxplot compares the distribution of child heights between males and females. We can see that male children tend to have a slightly higher median height than female children, although the distributions overlap considerably. Both groups also show a similar spread in heights. This suggests that while gender may play a role in determining height, it is likely only one of several factors influencing a child's height.

```{r}
ggplot(GaltonFamilies, aes(x = mother, y = father)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "red") +
  labs(
    title = "Mother vs Father Height",
    x = "Mother Height",
    y = "Father Height"
  )
```
This scatterplot shows the relationship between the heights of mothers and fathers in the dataset. There appears to be a slight upward trend, suggesting that taller fathers tend to have somewhat taller partners as well. However, the points are still fairly spread out, so the relationship is not very strong. Because the two variables are somewhat related, using both mother and father height as predictors in a regression model could introduce a small amount of multicollinearity, meaning the variables may contain overlapping information about parental height.

## Analysis

```{r}

# Split the data into training (80%) and testing (20%) sets
galton_split <- initial_split(GaltonFamilies, prop = 0.8)

galton_train <- training(galton_split)
galton_test  <- testing(galton_split)
```

```{r}
# Specify the linear regression model
lm_spec <- linear_reg() |>
  set_engine("lm") |>
  set_mode("regression")

# Define the model formula 
lm_recipe <- recipe(
  child_height ~ midparent_height + gender,
  data = galton_train)

# Build workflow and fit the model
lm_fit <- workflow() |>
  add_recipe(lm_recipe) |>
  add_model(lm_spec) |>
  fit(data = galton_train)

# View model summary
lm_fit

# Extract the regression coefficients
coeffs <- lm_fit |>
             extract_fit_parsnip() |>
             tidy()

# View coefficients
coeffs
```

```{r}
# Predict on the test data and evaluate model performance
lm_test_results <- lm_fit |>
  predict(galton_test) |>
  bind_cols(galton_test) |>
  metrics(truth = child_height, estimate = .pred)

# View test results
lm_test_results

# Scatterplot with regression line
ggplot(galton_heights, aes(x = midparent_height, y = child_height, color = gender)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    x = "Midparent Height (inches)",
    y = "Child Height (inches)",
    title = "Graph 1. Relationship Between Midparent Height and Child Height")
```

## Results

The fitted regression model produced the following equation:

$$
\text{Child Height} = 16.14 + 0.693(\text{Midparent Height}) + 5.21(\text{Male})
$$ 

The results indicate that midparent height is a significant predictor of child height (β = 0.693, p \< 0.001). For every one inch increase in midparent height, the predicted child height increases by approximately 0.69 inches.

Child gender was also a significant predictor (β = 5.21, p \< 0.001). Male children were predicted to be approximately 5.21 inches taller than female children on average, when holding the midparent height constant.

The model achieved a root mean squared error (RMSE) of 2.28 inches and a mean absolute error (MAE) of 1.76 inches, indicating that predictions were typically within approximately 2 inches of the true child height. The model explained 61.6% of the variation in child height (R² = 0.616).

Graph 1 shows the relationship between midparent height and child height, with separate regression lines for male and female children. The positive slope for both lines indicates that taller parents tend to have taller children. The colour separation between male and female points in the scatterplot reflects the gender difference, with male children generally having greater predicted heights than female children at the same midparent height.

## Discussion

## References
Galton, F. (1886). Regression Towards Mediocrity in Hereditary Stature **Journal of the Anthropological Institute**, 15, 246-263